Spatial Generalized Linear Models in \proglang{R} Using \pkg{spmodel}

### 2025-08-21 inst/manuscript/manuscript.Rmd
update to rmd from qmd

---
title: "Spatial Generalized Linear Models in [R]{.proglang} Using [spmodel]{.pkg}"
format:
    jss-pdf:
        keep-tex: true
    jss-html: default
author:
  - name: Michael Dumelle
    affiliations:
      - name: |
          | United States
          | Environmental Protection Agency
        department: Pacific Ecological Systems Division
        address: 200 SW 35th St
        city: Corvallis
        country: OR
        postal-code: 97333
    orcid: 0000-0002-3393-5529
    email: dumelle.michael@epa.gov
  - name: Jay M. Ver Hoef
    affiliations:
      - name: |
          | National Oceanic and
          | Atmospheric Administration
        department: Alaska Fisheries Science Center
    orcid: 0000-0003-4302-6895
  - name: Matt Higham
    affiliations:
      - name: St. Lawrence University
        department: Department of Math, Computer Science, and Statistics
    orcid: 0009-0006-4217-625X
abstract: |
  Generalized linear models (GLMs) describe a non-normal response variable that may be binary, count, skewed, or a proportion. Typically, observations in a GLM are assumed independent of one another. For spatial data, this independence assumption is impractical, as nearby locations tend to be more similar than locations far apart. The [spmodel]{.pkg} [R]{.proglang} package provides tools to fit GLMs that incorporate spatial autocorrelation (i.e., spatial generalized linear models, or SPGLMs). SPGLMs are fit in [spmodel]{.pkg} using a novel application of the Laplace approximation via [spglm]{.fct} for point-referenced data or [spgautor]{.fct} for areal (i.e., lattice), data. [spglm]{.fct} and [spgautor]{.fct} closely resemble [glm]{.fct} from base [R]{.proglang} but include arguments that control the spatial autocorrelation structure. [spmodel]{.pkg} has many helper functions for model inspection and diagnostics, some of which leverage other [R]{.proglang} packages like [broom]{.pkg} and [emmeans]{.pkg}. [spmodel]{.pkg} has tools to make predictions of the latent spatial-mean process at unobserved locations. [spmodel]{.pkg} also provides many advanced features like accommodating geometric anisotropy and nonspatial random effects, simulating spatially autocorrelated data, and more. Here we use [spmodel]{.pkg} to illustrate the modeling of binary, count, skewed and proportion response variables from several point-referenced and areal data sets. 
keywords: [Autoregressive Model, Geostatistics, Spatial Covariance, Spatially-Explicit Model, Statistical Model]
keywords-formatted: [Autoregressive Model, Geostatistics, Spatial Covariance, Spatially-Explicit Model, Statistical Model]
bibliography: references.bib  
preamble: |
  \usepackage{amsmath,amsfonts,amssymb}
  \usepackage{bm, bbm}
  \usepackage{lineno}
  \usepackage{caption, subcaption}
editor_options: 
  chunk_output_type: console
---

```{R}
#| include: false
library(spmodel)
library(ggplot2)
library(dplyr)
library(emmeans)
library(here)
fig_path <- here("inst", "manuscript", "figures")
```

@@

## Introduction {#sec-intro}

@@

The rest of this article is organized as follows. In @sec-spglm, we provide some background for the SPGLM fitting and prediction routines in [spmodel]{.pkg}. In @sec-applications, we provide several applications of [spmodel]{.pkg} to spatial binary, count, skewed and proportion data with both point-referenced and areal supports. And in @sec-discussion, we end with a discussion synthesizing \pkg{spmodel}'s contributions to the analysis of SPGLMs in [R]{.proglang}.

## Spatial generalized linear models using the Laplace approximation {#sec-spglm}

@@

## Application {#sec-applications}

The [spglm]{.fct} (for point-referenced data) and [spgautor]{.fct} (for areal data) functions in [spmodel]{.pkg} fit SPGLMs using the Laplace approximation outlined in @sec-spglm. Both [spglm]{.fct} and [spgautor]{.fct} generally require the following four arguments: `formula`, a formula that describes the relationship between the response variable and explanatory variables; `family`, the response distribution (which can be `binomial`, `poisson`, `nbinomial`, `Gamma`, `inverse.gaussian`, or `beta`); `data`, the data frame that holds the variables in `formula` as well as spatial locations; and `spcov_type`, the spatial covariance type. The first three arguments are shared by [glm]{.fct}; thus, the transition from GLMs to SPGLMs requires only one additional argument: `spcov_type`. The [spglm]{.fct} spatial covariance types measure dependence as a function of Euclidean distance among observations; an example is the exponential spatial covariance:
\begin{equation}\label{eq-spcov-exp}
  \boldsymbol{\Sigma} = \sigma^2_{de} \exp(-\mathbf{H}/\phi) + \sigma^2_{ie}\mathbf{I},
\end{equation}
where $\mathbf{H}$ is a matrix of pairwise distances among all observations. [spglm]{.fct} currently supports 18 distinct spatial covariance functions.

@@

In the rest of this section, we use [spmodel]{.pkg} to study binary, count, skewed, and proportion response variables that are either point-referenced or areal. We use [spmodel]{.pkg} for all parts of the data analysis, from estimation to inference to model diagnostics to prediction. We first describe core [spmodel]{.pkg} functionality in an application to binary data, while additional analyses highlight count, skewed, and proportion data as well as some additional [spmodel]{.pkg} features. Before proceeding, load [spmodel]{.pkg} into the current [R]{.proglang} session:
```{R}
#| prompt: true
#| eval: false
library("spmodel")
```

### Binary data

@@

```{R}
#| prompt: true
head(moose)
```

There are five columns: `elev`, the numeric site elevation (meters); `strat` a stratification variable for sampling with two levels, `"L"` and `"M"`, which are categorized by landscape metrics at each site; `count`, the number of moose at each site; `presence`, a factor that indicates whether at least one moose was observed at each site (`0` implies no moose; `1` implies at least one moose); and `geometry`, the NAD83 projected coordinate of each site. The `moose_preds` data in [spmodel]{.pkg} contain spatial locations at which predictions of moose presence are desired. They are also an `sf` object with the same projection and measurements for `elev` and `strat`. @fig-moose-data shows the `presence` variable in `moose` as well as the spatial locations of both `moose` and `moose_preds`. Moose are most common in the southwestern and eastern parts of the domain and least common in the northwest.
```{R}
#| prompt: false
#| echo: false
#| fig-align: center
#| fig-cap: "Moose presence in Alaska. Circles represent moose presence or absence (based on color) and triangles represent locations at which moose presence probability predictions are desired."
#| out-width: "70%"
#| out-height: "70%"
#| label: fig-moose-data
knitr::include_graphics(paste0(fig_path, "/figure-1.png"))
```

@@

```{R}
#| prompt: true
spbin <- spglm(
  formula = presence ~ elev + strat + elev:strat,
  family = binomial,
  data = moose,
  spcov_type = "spherical"
)
```

Summarizing the model object yields a summary similar to that provided by the familiar [glm]{.fct}:
```{R}
#| prompt: true
summary(spbin)
```

The summary contains the original function call, a summary of residuals, the fixed effects coefficients table, the spatial covariance parameter estimates, and additional model information like the pseudo R-squared, which quantifies the variability in the model attributable to the fixed effects. While useful, this summary information is hard to work with, as it is printed directly to the [R]{.proglang} console. The [broom]{.pkg} package from the tidymodels [@kuhn2022tidy] ecosystem has functions to provide helpful model output in the form of tibbles (i.e., data frames) that are easily manipulated. [spmodel]{.pkg} has methods for the [tidy]{.fct}, [glance]{.fct}, and [augment]{.fct} functions from [broom]{.pkg}. The first [broom]{.pkg} function is [tidy]{.fct}, which tidies the model output:
```{R}
#| prompt: true
tidy(spbin)
```

The estimates and standard errors returned are on the log odds link (Table$~$\ref{table-links}) scale ([coef]{.fct} and [vcov]{.fct} may also be used). The output provides evidence that elevation is positively associated with moose presence in the `"L"` stratum ($p-$value $<0.05$) and, at zero elevation, moose are more likely in the `"M"` stratum than the `"L"` stratum ($p-$value $<0.01$). This output provides marginal evidence that the effect of elevation on moose presence varies across strata ($p-$value $\approx 0.1$).  The model effectively quantifies the impact of elevation on moose presence for moose in the `"L"` strata, but an analogous statement for moose in the `"M"` strata requires more context. We could refit the model treating `"M"` as the reference group instead of `"L"`:
```{R}
#| prompt: true
moose$strat2 <- factor(moose$strat, levels = c("M", "L"))
update(spbin, formula = presence ~ elev + strat2 + elev:strat2) |> 
  summary()
```

@@

```{r}
#| prompt: true
#| eval: false
library("emmeans")
```

```{R}
#| prompt: true
emtrends(spbin, "strat", "elev")
```

@@

The second [broom]{.pkg} function is [glance]{.fct}, which glances at the model fit:
```{R}
#| prompt: true
glance(spbin)
```

@@

```{R}
#| prompt: false
#| echo: false
#| label: fig-sp-diagnostic
#| fig-cap: "Spatial logistic regression model diagnostics from [augment]{.fct}. The leverage (i.e., hat) values (left) and standardized residuals (right)."
#| out-width: "100%"
#| out-height: "100%"
knitr::include_graphics(paste0(fig_path, "/figure-2.png"))
```

@@

The [augment]{.fct} function returns the fitted $\mathbf{w}$ values (`.fitted`), deviance residuals (`.resid`), leverage (i.e., hat) values (`.hat`), Cook's distance (`.cooksd`), and standardized residuals (`.std.resid`). When the data are an `sf` object, [augment]{.fct} returns another `sf` object, helpful for visualizing model diagnostics spatially as in @fig-sp-diagnostic. Leverage measures the unusualness of an observation's set of explanatory variables, while Cook's distances measures how influential an observation is on the resulting model fit [@montgomery2021introduction]. Model diagnostics are also accessible as vectors using the appropriate generic function (e.g., [fitted]{.fct}, [residuals]{.fct}).

Similar to [glm]{.fct} model objects, [plot]{.fct} can be used to visualize diagnostics (@fig-sp-diagnostic2):
```{R}
#| prompt: true
#| eval: false
plot(spbin, which = c(4, 7))
```

```{R}
#| prompt: true
#| echo: false
#| label: fig-sp-diagnostic2
#| fig-cap: "Spatial logistic regression model diagnostics from [plot]{.fct}. The Cook's distance values (left) and the fitted spatial covariance as a function of distance (right)."
#| out-width: "100%"
#| out-height: "100%"
knitr::include_graphics(paste0(fig_path, "/figure-3.png"))
```

Components of model variation are partitioned using [varcomp]{.fct}:
```{R}
#| prompt: true
varcomp(spbin)
```
The fixed effects explain roughly 9% of model variation, while the spatially dependent variance explains most of the remaining variability.

```{R}
#| prompt: false
#| echo: false
#| fig-align: center
#| fig-cap: "Moose presence probability fitted values and predictions. Fitted values are represeneted by circles and predictions by triangles."
#| out-width: "70%"
#| out-height: "70%"
#| label: fig-moose-fit
knitr::include_graphics(paste0(fig_path, "/figure-4.png"))
```

We make predictions of the log odds of moose probability presence at each site in `moose_preds` using [predict]{.fct}:
```{R}
#| prompt: true
head(predict(spbin, newdata = moose_preds))
```

[augment]{.fct} may also be used to augment the prediction data with predictions:
```{R}
head(augment(
  spbin,
  newdata = moose_preds,
  type.predict = "response",
  interval = "prediction"
))
```

Here, we requested predictions on the probability (i.e., response) scale (@fig-moose-fit) alongside lower and upper bounds of a 95% (see `level`) prediction interval (@fig-moose-int).

```{R}
#| prompt: false
#| echo: false
#| label: fig-moose-int
#| fig-cap: "Moose presence probability prediction intervals. 95% prediction interval lower bound (left) and 95% prediction interval upper bound (right)."
#| out-width: "100%"
#| out-height: "100%"
knitr::include_graphics(paste0(fig_path, "/figure-5.png"))
```

@@

```{R}
#| prompt: true
# fit nonspatial model using spglm()
bin <- spglm(
  formula = presence ~ elev + strat + elev:strat,
  family = binomial,
  data = moose,
  spcov_type = "none"
)
# fit nonspatial model using glm()
bin_glm <- glm(
  formula = presence ~ elev + strat + elev:strat,
  family = binomial,
  data = moose
)
# compare fixed effect coefficients and standard errors up to four decimals
data.frame(
  est_none = coef(bin),
  est_glm = coef(bin_glm),
  se_none = sqrt(diag(vcov(bin))),
  se_glm = sqrt(diag(vcov(bin_glm)))
) |>
  apply(2, round, digits = 4)
```

The advantage of using [spglm]{.fct} to fit a model with `spcov_type = "none"` is that it provides access to other [spmodel]{.pkg} functions for model objects (e.g., [glances]{.fct} below) and accounts for the additional terms in the likelihood from Equation$~$\eqref{eq-marginal6}. These additional terms make the likelihood for [spglm]{.fct} and [glm]{.fct} different, though the models convey the same information. A glance at the spatial and nonspatial models reveals:
```{R}
#| prompt: true
glances(spbin, bin)
```

The spatial model has a notably lower AIC, AICc, BIC, and deviance, suggesting it is the superior model. Another model comparison approach is leave-one-out cross validation. In leave-one-out cross validation, separately each observation is held out, a model is fit to the remaining data, and a prediction is made for the mean of the held out observation on the response scale. Then, statistics like leave-one-out bias, mean-squared-prediction error (MSPE), and the square root of MSPE (RMSPE) may be computed:
```{R}
#| prompt: true
loocv(spbin) |>
  apply(2, round, digits = 4)
loocv(bin) |>
  apply(2, round, digits = 4)
```

@@

A third model comparison tool is area under the receiver operating characteristic (AUROC) curve. The AUROC curve ranges from zero to one and conveys a model's classification performance over all possible probability thresholds [@james2013introduction]. Larger values of AUROC indicate a more accurate model:
```{R}
#| prompt: true
AUROC(spbin)
AUROC(bin)
```

@@

### Count data

```{R}
#| prompt: false
#| echo: false
#| fig-align: center
#| fig-cap: "Moose counts in Alaska. Circles represent moose counts (based on color) and triangles represent locations at which mean count predictions are desired."
#| label: fig-moose-data-count
#| out-width: "70%"
#| out-height: "70%"
knitr::include_graphics(paste0(fig_path, "/figure-6.png"))
```

The `count` variable in `moose` contains the number of moose observed at a site (@fig-moose-data-count). Count data are often modeled using Poisson or negative binomial regression with the log link function. The Poisson regression model assumes each datum's underlying latent mean equals its variance, while the negative binomial accommodates overdispersion (where the variance is greater than the mean) at the cost of estimating an extra parameter.

@@

```{r}
#| prompt: true
sppois <- spglm(
  formula = count ~ elev + strat + elev:strat,
  family = poisson,
  data = moose,
  spcov_type = "gaussian",
  anisotropy = TRUE
)

@@

Because the models have the same support (i.e., both non-negative count models), we can use likelihood-based statistics to compare them:
```{r}
#| prompt: true
glances(sppois, spnbin, sort_by = "AIC") |>
  subset(select = c(model, npar, AIC, AICc, BIC))
```

The negative binomial model has a slightly lower AIC and AICc, while the Poisson model has a slightly lower BIC. This is reasonable given the BIC penalizes additional parameters (here, an overdispersion parameter) more heavily than AIC and AICc. The leave-out-out MSPE prefers the negative binomial model:
```{R}
#| prompt: true
loocv(sppois) |>
  apply(2, round, digits = 4)
loocv(spnbin) |>
  apply(2, round, digits = 4)
```

A likelihood-based comparison between the negative binomial anisotropic model and the negative binomial isotropic model suggests that the anisotropic model is preferred (lower AIC, AICc, and BIC):
```{R}
#| prompt: true
spnbin_iso <- update(spnbin, anisotropy = FALSE)
glances(spnbin_iso, spnbin, sort_by = "AIC") |>
  subset(select = c(model, npar, AIC, AICc, BIC))
```

[plot]{.fct} returns the spatial covariance as a function of direction (@fig-tropy):
```{R}
#| prompt: true
#| eval: false
plot(spnbin_iso, which = 8)
plot(spnbin, which = 8)
```

```{R}
#| prompt: true
#| echo: false
#| label: fig-tropy
#| fig-cap: "Level curves of equal autocorrelation for the negative binomial moose count models. The ellipse is centered at zero distance in the x-direction and y-direction, and points along the ellipse have equal levels of autocorrelation.  In the isotropic level curve (left), spatial covariance decays equally in all directions. In the anistropic level curve (right), spatial covariance decays fastest in the northeast-southwest direction and slowest in the northwest-southeast direction (this pattern can be seen in the observed counts)."
#| out-width: "100%"
#| out-height: "100%"
#| fig-asp: 1
knitr::include_graphics(paste0(fig_path, "/figure-7.png"))
```

Earlier we used [tidy]{.fct} to tidy the model's fixed effects, but we can also use tidy to tidy the spatial covariance parameters:
```{R}
#| prompt: true
tidy(spnbin, effects = "spcov")
```

@@

### Skewed data

@@

The trend data were originally logged to remove skew [@ver2018spatial], but we will exponentiate `log_trend` (@fig-seal) and model this skew directly using a SPGLM:
```{R}
#| prompt: true
seal$trend <- exp(seal$log_trend)
```

@@

```{r}
#| prompt: true
#| echo: false
#| label: fig-seal
#| fig-cap: "Seal trend distribution in Alaska. Observed and missing seal polygons by stock (left) and observed seal trends (right)."
#| out-width: "100%"
#| out-height: "100%"
knitr::include_graphics(paste0(fig_path, "/figure-8.png"))
```

[spmodel]{.pkg} supports the gamma and inverse Gaussian families for modeling skewed, positive response variables. [spmodel]{.pkg} also supports nonspatial random effects specified via the `random` argument, which uses a similar formula syntax as [nlme]{.pkg} [@pinheiro2006mixed] and [lme4]{.pkg} [@bates2015lme4]. Using likelihood-based statistics, we compare two models fit using the simultaneous autoregressive covariance and the Gamma and inverse Gaussian families. Both models have a random effect for seal stock, which builds additional correlation into the model for two polygons from the same stock:
```{r}
#| prompt: true
spgam <- spgautor(
  formula = trend ~ 1,
  family = Gamma,
  data = seal,
  spcov_type = "sar",
  random = ~ stock
)
spinvg <- update(spgam, family = inverse.gaussian)
glances(spgam, spinvg, sort_by = "AIC") |>
  subset(select = c(model, npar, AIC, AICc, BIC))
```

@@

## Discussion {#sec-discussion}

@@

## Computational details {.unnumbered}

>>>>

---
documentclass: jss
author:
    # use this syntax to add text on several lines
    # To add another line, use \AND at the end of the previous one as above
    # use a different affiliation in adress field (differently formated here)
  - name: Michael Dumelle
    orcid: 0000-0002-3393-5529
    affiliation: |
      | United States
      | Environmental Protection Agency
    address: |
      | 200 SW 35th St
      | Corvallis, OR, 97330
    email: \email{Dumelle.Michael@epa.gov}
  - name: Jay M. Ver Hoef
    orcid: 0000-0003-4302-6895
    affiliation: |
      | Alaska Fisheries
      | Science Center
  - name: Matt Higham
    orcid: 0009-0006-4217-625X
    affiliation: |
      | St. Lawrence University
title:
  # If you use tex in the formatted title, also supply version without
  # For running headers, if needed
  formatted: "Spatial Generalized Linear Models in [R]{.proglang} Using [spmodel]{.pkg}"
  plain:     "Spatial Generalized Linear Models in R Using spmodel"
  short:     "SPGLMs in [R]{.proglang} Using [spmodel]{.pkg}"
abstract: |
  Generalized linear models (GLMs) describe a non-normal response variable that may be binary, count, skewed, or a proportion. Typically, observations in a GLM are assumed independent of one another. For spatial data, this independence assumption is impractical, as nearby locations tend to be more similar than locations far apart. The [spmodel]{.pkg} [R]{.proglang} package provides tools to fit GLMs that incorporate spatial autocorrelation (i.e., spatial generalized linear models, or SPGLMs). SPGLMs are fit in [spmodel]{.pkg} using a novel application of the Laplace approximation via [spglm]{.fct} for point-referenced data or [spgautor]{.fct} for areal (i.e., lattice), data. [spglm]{.fct} and [spgautor]{.fct} closely resemble [glm]{.fct} from base [R]{.proglang} but include arguments that control the spatial autocorrelation structure. [spmodel]{.pkg} has many helper functions for model inspection and diagnostics, some of which leverage other [R]{.proglang} packages like [broom]{.pkg} and [emmeans]{.pkg}. [spmodel]{.pkg} has tools to make predictions of the latent spatial-mean process at unobserved locations. [spmodel]{.pkg} also provides many advanced features like accommodating geometric anisotropy and nonspatial random effects, simulating spatially autocorrelated data, and more. Here we use [spmodel]{.pkg} to illustrate the modeling of binary, count, skewed and proportion response variables from several point-referenced and areal data sets. 
keywords:
  # at least one keyword must be supplied
  formatted: [autoregressive model, geostatistical model, spatial autocovariance, spatial autocorrelation]
  plain:     [autoregressive model, geostatistical model, spatial autocovariance, spatial autocorrelation]
preamble: |
  \usepackage{amsmath,amsfonts,amssymb}
  \usepackage{bm, bbm}
  \usepackage{lineno}
  \usepackage{caption, subcaption}
output: rticles::jss_article
editor_options: 
  chunk_output_type: console
---

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
library(spmodel)
library(ggplot2)
library(dplyr)
library(emmeans)
library(here)
fig_path <- here("inst", "manuscript", "figures")
```

@@

# Introduction {#sec-intro}

@@

The rest of this article is organized as follows. In Section$~$\ref{sec-spglm}, we provide some background for the SPGLM fitting and prediction routines in [spmodel]{.pkg}. In Section$~$\ref{sec-applications}, we provide several applications of [spmodel]{.pkg} to spatial binary, count, skewed and proportion data with both point-referenced and areal supports. And in Section$~$\ref{sec-discussion}, we end with a discussion synthesizing \pkg{spmodel}'s contributions to the analysis of SPGLMs in [R]{.proglang}.

# Spatial generalized linear models using the Laplace approximation {#sec-spglm}

@@

# Application {#sec-applications}

The [spglm]{.fct} (for point-referenced data) and [spgautor]{.fct} (for areal data) functions in [spmodel]{.pkg} fit SPGLMs using the Laplace approximation outlined in Section$~$\ref{sec-spglm}. Both [spglm]{.fct} and [spgautor]{.fct} generally require the following four arguments: `formula`, a formula that describes the relationship between the response variable and explanatory variables; `family`, the response distribution (which can be `binomial`, `poisson`, `nbinomial`, `Gamma`, `inverse.gaussian`, or `beta`); `data`, the data frame that holds the variables in `formula` as well as spatial locations; and `spcov_type`, the spatial covariance type. The first three arguments are shared by [glm]{.fct}; thus, the transition from GLMs to SPGLMs requires only one additional argument: `spcov_type`. The [spglm]{.fct} spatial covariance types measure dependence as a function of Euclidean distance among observations; an example is the exponential spatial covariance:
\begin{equation}\label{eq-spcov-exp}
  \boldsymbol{\Sigma} = \sigma^2_{de} \exp(-\mathbf{H}/\phi) + \sigma^2_{ie}\mathbf{I},
\end{equation}
where $\mathbf{H}$ is a matrix of pairwise distances among all observations. [spglm]{.fct} currently supports 18 distinct spatial covariance functions.

@@

In the rest of this section, we use [spmodel]{.pkg} to study binary, count, skewed, and proportion response variables that are either point-referenced or areal. We use [spmodel]{.pkg} for all parts of the data analysis, from estimation to inference to model diagnostics to prediction. We first describe core [spmodel]{.pkg} functionality in an application to binary data, while additional analyses highlight count, skewed, and proportion data as well as some additional [spmodel]{.pkg} features. Before proceeding, load [spmodel]{.pkg} into the current [R]{.proglang} session:
```{r, eval = FALSE}
library("spmodel")
```

## Binary data

@@

```{r}
head(moose)
```

There are five columns: `elev`, the numeric site elevation (meters); `strat` a stratification variable for sampling with two levels, `"L"` and `"M"`, which are categorized by landscape metrics at each site; `count`, the number of moose at each site; `presence`, a factor that indicates whether at least one moose was observed at each site (`0` implies no moose; `1` implies at least one moose); and `geometry`, the NAD83 projected coordinate of each site. The `moose_preds` data in [spmodel]{.pkg} contain spatial locations at which predictions of moose presence are desired. They are also an `sf` object with the same projection and measurements for `elev` and `strat`. Figure$~$\ref{fig-moose-data} shows the `presence` variable in `moose` as well as the spatial locations of both `moose` and `moose_preds`. Moose are most common in the southwestern and eastern parts of the domain and least common in the northwest.

\begin{figure}
\centering
\includegraphics[width = 0.70\linewidth]{figures/figure-1.png}
\caption{Moose presence in Alaska. Circles represent moose presence or absence (based on color) and triangles represent locations at which moose presence probability predictions are desired.}
\label{fig-moose-data}
\end{figure}

@@

```{r}
spbin <- spglm(
  formula = presence ~ elev + strat + elev:strat,
  family = binomial,
  data = moose,
  spcov_type = "spherical"
)
```

Summarizing the model object yields a summary similar to that provided by the familiar [glm]{.fct}:
```{r}
summary(spbin)
```

The summary contains the original function call, a summary of residuals, the fixed effects coefficients table, the spatial covariance parameter estimates, and additional model information like the pseudo R-squared, which quantifies the variability in the model attributable to the fixed effects. While useful, this summary information is hard to work with, as it is printed directly to the [R]{.proglang} console. The [broom]{.pkg} package from the tidymodels [@kuhn2022tidy] ecosystem has functions to provide helpful model output in the form of tibbles (i.e., data frames) that are easily manipulated. [spmodel]{.pkg} has methods for the [tidy]{.fct}, [glance]{.fct}, and [augment]{.fct} functions from [broom]{.pkg}. The first [broom]{.pkg} function is [tidy]{.fct}, which tidies the model output:
```{r}
tidy(spbin)
```

The estimates and standard errors returned are on the log odds link (Table$~$\ref{table-links}) scale ([coef]{.fct} and [vcov]{.fct} may also be used). The output provides evidence that elevation is positively associated with moose presence in the `"L"` stratum ($p-$value $<0.05$) and, at zero elevation, moose are more likely in the `"M"` stratum than the `"L"` stratum ($p-$value $<0.01$). This output provides marginal evidence that the effect of elevation on moose presence varies across strata ($p-$value $\approx 0.1$).  The model effectively quantifies the impact of elevation on moose presence for moose in the `"L"` strata, but an analogous statement for moose in the `"M"` strata requires more context. We could refit the model treating `"M"` as the reference group instead of `"L"`:
```{r}
moose$strat2 <- factor(moose$strat, levels = c("M", "L"))
update(spbin, formula = presence ~ elev + strat2 + elev:strat2) |> 
  summary()
```

@@

```{r, eval = FALSE}
library("emmeans")
```

```{r}
emtrends(spbin, "strat", "elev")
```

@@

The second [broom]{.pkg} function is [glance]{.fct}, which glances at the model fit:
```{r}
glance(spbin)
```

@@

\begin{figure}
\centering
\includegraphics[width = 1\linewidth]{figures/figure-2.png}
\caption{Spatial logistic regression model diagnostics from [augment]{.fct}. The leverage (i.e., hat) values (left) and standardized residuals (right).}
\label{fig-sp-diagnostic}
\end{figure}

@@

The [augment]{.fct} function returns the fitted $\mathbf{w}$ values (`.fitted`), deviance residuals (`.resid`), leverage (i.e., hat) values (`.hat`), Cook's distance (`.cooksd`), and standardized residuals (`.std.resid`). When the data are an `sf` object, [augment]{.fct} returns another `sf` object, helpful for visualizing model diagnostics spatially as in Figure$~$\ref{fig-sp-diagnostic}. Leverage measures the unusualness of an observation's set of explanatory variables, while Cook's distances measures how influential an observation is on the resulting model fit [@montgomery2021introduction]. Model diagnostics are also accessible as vectors using the appropriate generic function (e.g., [fitted]{.fct}, [residuals]{.fct}).

Similar to [glm]{.fct} model objects, [plot]{.fct} can be used to visualize diagnostics (Figure$~$\ref{fig-sp-diagnostic2}):
```{r, eval = FALSE}
plot(spbin, which = c(4, 7))
```

\begin{figure}
\centering
\includegraphics[width = 1\linewidth]{figures/figure-3.png}
\caption{Spatial logistic regression model diagnostics from [plot]{.fct}. The Cook's distance values (left) and the fitted spatial covariance as a function of distance (right).}
\label{fig-sp-diagnostic2}
\end{figure}

Components of model variation are partitioned using [varcomp]{.fct}:
```{r}
varcomp(spbin)
```
The fixed effects explain roughly 9% of model variation, while the spatially dependent variance explains most of the remaining variability.

\begin{figure}
\centering
\includegraphics[width = 0.70\linewidth]{figures/figure-4.png}
\caption{Moose presence probability fitted values and predictions. Fitted values are represeneted by circles and predictions by triangles.}
\label{fig-moose-fit}
\end{figure}

We make predictions of the log odds of moose probability presence at each site in `moose_preds` using [predict]{.fct}:
```{r}
head(predict(spbin, newdata = moose_preds))
```

[augment]{.fct} may also be used to augment the prediction data with predictions:
```{r}
head(augment(
  spbin,
  newdata = moose_preds,
  type.predict = "response",
  interval = "prediction"
))
```

Here, we requested predictions on the probability (i.e., response) scale (Figure$~$\ref{fig-moose-fit}) alongside lower and upper bounds of a 95% (see `level`) prediction interval (Figure$~$\ref{fig-moose-int}).

\begin{figure}
\centering
\includegraphics[width = 1\linewidth]{figures/figure-5.png}
\caption{Moose presence probability prediction intervals. 95\% prediction interval lower bound (left) and 95\% prediction interval upper bound (right).}
\label{fig-moose-int}
\end{figure}

@@

```{r}
bin <- spglm(
  formula = presence ~ elev + strat + elev:strat,
  family = binomial,
  data = moose,
  spcov_type = "none"
)
bin_glm <- glm(
  formula = presence ~ elev + strat + elev:strat,
  family = binomial,
  data = moose
)
data.frame(
  est_none = coef(bin),
  est_glm = coef(bin_glm),
  se_none = sqrt(diag(vcov(bin))),
  se_glm = sqrt(diag(vcov(bin_glm)))
) |>
  apply(2, round, digits = 4)
```

The advantage of using [spglm]{.fct} to fit a model with `spcov_type = "none"` is that it provides access to other [spmodel]{.pkg} functions for model objects (e.g., [glances]{.fct} below) and accounts for the additional terms in the likelihood from Equation$~$\eqref{eq-marginal6}. These additional terms make the likelihood for [spglm]{.fct} and [glm]{.fct} different, though the models convey the same information. A glance at the spatial and nonspatial models reveals:
```{r}
glances(spbin, bin)
```

The spatial model has a notably lower AIC, AICc, BIC, and deviance, suggesting it is the superior model. Another model comparison approach is leave-one-out cross validation. In leave-one-out cross validation, separately each observation is held out, a model is fit to the remaining data, and a prediction is made for the mean of the held out observation on the response scale. Then, statistics like leave-one-out bias, mean-squared-prediction error (MSPE), and the square root of MSPE (RMSPE) may be computed:
```{r}
loocv(spbin) |>
  apply(2, round, digits = 4)
loocv(bin) |>
  apply(2, round, digits = 4)
```

@@

A third model comparison tool is area under the receiver operating characteristic (AUROC) curve. The AUROC curve ranges from zero to one and conveys a model's classification performance over all possible probability thresholds [@james2013introduction]. Larger values of AUROC indicate a more accurate model:
```{r}
AUROC(spbin)
AUROC(bin)
```

@@

## Count data

\begin{figure}
\centering
\includegraphics[width = 0.70\linewidth]{figures/figure-6.png}
\caption{Moose counts in Alaska. Circles represent moose counts (based on color) and triangles represent locations at which mean count predictions are desired.}
\label{fig-moose-data-count}
\end{figure}

The `count` variable in `moose` contains the number of moose observed at a site (Figure$~$\ref{fig-moose-data-count}). Count data are often modeled using Poisson or negative binomial regression with the log link function. The Poisson regression model assumes each datum's underlying latent mean equals its variance, while the negative binomial accommodates overdispersion (where the variance is greater than the mean) at the cost of estimating an extra parameter.

@@

```{r}
sppois <- spglm(
  formula = count ~ elev + strat + elev:strat,
  family = poisson,
  data = moose,
  spcov_type = "gaussian",
  anisotropy = TRUE
)

@@

Because the models have the same support (i.e., both non-negative count models), we can use likelihood-based statistics to compare them:
```{r}
glances(sppois, spnbin, sort_by = "AIC") |>
  subset(select = c(model, npar, AIC, AICc, BIC))
```

The negative binomial model has a slightly lower AIC and AICc, while the Poisson model has a slightly lower BIC. This is reasonable given the BIC penalizes additional parameters (here, an overdispersion parameter) more heavily than AIC and AICc. The leave-out-out MSPE prefers the negative binomial model:
```{r}
loocv(sppois) |>
  apply(2, round, digits = 4)
loocv(spnbin) |>
  apply(2, round, digits = 4)
```

A likelihood-based comparison between the negative binomial anisotropic model and the negative binomial isotropic model suggests that the anisotropic model is preferred (lower AIC, AICc, and BIC):
```{r}
spnbin_iso <- update(spnbin, anisotropy = FALSE)
glances(spnbin_iso, spnbin, sort_by = "AIC") |>
  subset(select = c(model, npar, AIC, AICc, BIC))
```

[plot]{.fct} returns the spatial covariance as a function of direction (Figure$~$\ref{fig-tropy}):
```{r}
plot(spnbin_iso, which = 8)
plot(spnbin, which = 8)
```

\begin{figure}
\centering
\includegraphics[width = 1\linewidth]{figures/figure-7.png}
\caption{Level curves of equal autocorrelation for the negative binomial moose count models. The ellipse is centered at zero distance in the x-direction and y-direction, and points along the ellipse have equal levels of autocorrelation.  In the isotropic level curve (left), spatial covariance decays equally in all directions. In the anistropic level curve (right), spatial covariance decays fastest in the northeast-southwest direction and slowest in the northwest-southeast direction (this pattern can be seen in the observed counts).}
\label{fig-tropy}
\end{figure}

Earlier we used [tidy]{.fct} to tidy the model's fixed effects, but we can also use tidy to tidy the spatial covariance parameters:
```{r}
tidy(spnbin, effects = "spcov")
```

@@

## Skewed data

@@

The trend data were originally logged to remove skew [@ver2018spatial], but we will exponentiate `log_trend` (Figure$~$\ref{fig-seal}) and model this skew directly using a SPGLM:
```{r}
seal$trend <- exp(seal$log_trend)
```

@@

\begin{figure}
\centering
\includegraphics[width = 1\linewidth]{figures/figure-8.png}
\caption{Seal trend distribution in Alaska. Observed and missing seal polygons by stock (left) and observed seal trends (right).}
\label{fig-seal}
\end{figure}

[spmodel]{.pkg} supports the gamma and inverse Gaussian families for modeling skewed, positive response variables. [spmodel]{.pkg} also supports nonspatial random effects specified via the `random` argument, which uses a similar formula syntax as [nlme]{.pkg} [@pinheiro2006mixed] and [lme4]{.pkg} [@bates2015lme4]. Using likelihood-based statistics, we compare two models fit using the simultaneous autoregressive covariance and the Gamma and inverse Gaussian families. Both models have a random effect for seal stock, which builds additional correlation into the model for two polygons from the same stock:
```{r}
spgam <- spgautor(
  formula = trend ~ 1,
  family = Gamma,
  data = seal,
  spcov_type = "sar",
  random = ~ stock
)
spinvg <- update(spgam, family = inverse.gaussian)
glances(spgam, spinvg, sort_by = "AIC") |>
  subset(select = c(model, npar, AIC, AICc, BIC))
```

@@

# Discussion {#sec-discussion}

@@

# Computational details {.unnumbered}
